---
title: "Management Strategy Evaluation with FLIBM"
author: "Marc H. Taylor"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: TRUE
    df_print: paged
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.align='center', comment=NA,
  message=FALSE, warning=FALSE, echo=TRUE, cache=FALSE,
  fig.width=6, fig.height=4.5)
```


# Introduction


# Setup
The FLIBM imports the package `FLCore` and recommends `ggplotFL` for use of summary plotting functions.

```{r}
# required packages
library(FLIBM)
library(FLCore)
library(ggplotFL)
library(data.table)
library(FLa4a)
library(FLXSA)
library(FLAssess)
library(FLash)
library(FLBRP)
```


```{r}
data(stk1) # load
stk <- stk1 # rename

# collapse season dimension and remove years with few to no catches
tmp <- simplify(stk$stock.a, dims = "season")

# exclude zero age class
stkTrue <- tmp[ac(1:range(tmp)["max"]),] 

# redefine plusgroup to collapse rare ages
stkTrue <- setPlusGroup(stkTrue, plusgroup = 4, na.rm=TRUE) 

# fill in missing slots
stkTrue@landings.n <- stkTrue@catch.n
stkTrue@landings.wt <- stkTrue@catch.wt
stkTrue@discards.n[] <- 0
stkTrue@discards.wt[] <- 0
stock(stkTrue) <- computeStock(stkTrue)
landings(stkTrue) <- computeLandings(stkTrue)
catch(stkTrue) <- computeCatch(stkTrue)
discards(stkTrue) <- computeDiscards(stkTrue)

# make assumption for natural mortality
stkTrue@m[] <- 0.7

# plot
plot(stkTrue)
```





# Management Strategy Evaluation
To come...

```{r}
it <- 20    # iterations
y0 <- range(stkTrue)["minyear"] # initial data year
dy <- range(stkTrue)["maxyear"] # final data year
iy <- dy+1  # initial year of projection (also intermediate year)
ny <- 12    # number of years to project from intial year
fy <- dy+ny # final year
nsqy <- 3   # number of years to compute status quo metrics


# Fit stock assessment model a4a
fmod <-  formula( ~ factor(age) + s(year, k=8) ) # fishing mortality submodel
srmod <- formula( ~ s(year, k=6)) # recruitment submodel
qmod <- list( ~ factor(age)) # survey catchability submodel

mcsave <- 100
mcmc <- it * mcsave

# Fit the model
fit <- sca(stkTrue, idx, 
  fmodel = fmod, qmodel = qmod, srmodel=srmod, fit = "MCMC",
  mcmc = SCAMCMC(mcmc = mcmc, mcsave = mcsave, mcprobe = 0.4))

# Update the FLStock object
stk_a4a <- stkTrue + fit

# Reduce to keep one iteration only for reference points
stk0 <- qapply(stk_a4a, iterMedians)
plot(stk_a4a)

```

```{r sr, message=FALSE, warning=FALSE, results="hide"}
# Fit the stock-recruit model
srbh <- fmle(as.FLSR(stk_a4a, model="bevholt"), method="L-BFGS-B", 
        lower=c(1e-6, 1e-6), upper=c(max(rec(stk_a4a)) * 3, Inf))
srbh0 <- fmle(as.FLSR(stk0, model="bevholt"), method="L-BFGS-B", 
         lower=c(1e-6, 1e-6), upper=c(max(rec(stk_a4a)) * 3, Inf))
# Generate stock-recruit residuals for the projection period
srbh.res <- rnorm(it, FLQuant(0, dimnames=list(year=iy:fy)), mean(c(apply(residuals(srbh), 6, sd))))

plot(srbh0) 
```




```{r, refpts}
# Calculate the reference points
brp <- brp(FLBRP(stk0, srbh0))
Fmsy <- c(refpts(brp)["msy","harvest"])
msy <- c(refpts(brp)["msy","yield"])
Bmsy <- c(refpts(brp)["msy","ssb"])
Bpa <- 0.5*Bmsy
Blim <- Bpa/1.4
plot(brp)
# Prepare the FLStock object for projections
stk_a4a <- stf(stk_a4a, fy-dy, nsqy, nsqy)
plot(stk_a4a)
```

# SET UP OBSERVATION ERROR MODEL ELEMENTS

**Estimate the index catchabilities from the a4a fit (without simulation)**

Observation error is introduced through the index catchability-at-age  

```{r, idx}
# Set up the FLIndices object and populate it
# (note, FLIndices potentially has more than one index, hence the for loop)
idcs <- FLIndices()
for (i in 1:length(idx)){
#   Set up FLQuants and calculate mean and sd for catchability
    lst <- mcf(list(index(idx), stock.n(stk0))) # make FLQuants same dimensions
    idx.lq <- log(lst[[1]]/lst[[2]]) # log catchability of index
    idx.qmu <- idx.qsig <- stock.n(iter(stk_a4a,1)) # create quants
    idx.qmu[] <- yearMeans(idx.lq) # allocate same mean-at-age to every year
    idx.qsig[] <- sqrt(yearVars(idx.lq)) # allocate same sd-at-age to every year
#   Build index catchability based on lognormal distribution with mean and sd calculated above
    idx.q <- rlnorm(it, idx.qmu, idx.qsig)
    idx_temp <- idx.q * stock.n(stk_a4a)
    idx_temp <- FLIndex(index=idx_temp, index.q=idx.q) # generate initial index
    range(idx_temp)[c("startf", "endf")] <- c(0, 0) # timing of index (as proportion of year)
    idcs[[i]] <- idx_temp
}
names(idcs) <- names(idx)
idx<-idcs[1]
```

# SET UP MSE LOOP

## Needed Functions  

**Observation error model**  
In this tutorial, observation error is applied to the operating model population numbers to obtain an index of abundance. This is implemented through the index catchability-at-age. Observation error is added during each year of the projection window, and is therefore dealt with more easily in a function.
```{r, oem}
o <- function(stk, idx, assessmentYear, dataYears) {
	# dataYears is a position vector, not the years themselves
	stk.tmp <- stk[, dataYears]
	# add small amount to avoid zeros
	catch.n(stk.tmp) <- catch.n(stk.tmp) + 0.1
	# Generate the indices - just data years
	idx.tmp <- lapply(idx, function(x) x[,dataYears])
  # Generate observed index
	for (i in 1:length(idx)) index(idx[[i]])[, assessmentYear] <- 
    stock.n(stk)[, assessmentYear]*index.q(idx[[i]])[, assessmentYear]
  list(stk=stk.tmp, idx=idx.tmp, idx.om=idx)
}
```

**XSA assessment model**  

The assessment model used to parameterise the HCR is XSA, through `FLXSA`. This function sets the control parameters for `FLXSA` and fits the assessment.

```{r, xsa}
xsa <- function(stk, idx){
  # Set up XSA settings
    control  <- FLXSA.control(tol = 1e-09, maxit=99, min.nse=0.3, fse=2.0,
                              rage = -1, qage = range(stk)["max"]-1, shk.n = TRUE, shk.f = TRUE,
                              shk.yrs = 5, shk.ages= 5, window = 100, tsrange = 99, tspower = 0)
  # Fit XSA
  fit <- FLXSA(stk, idx, control)
  # convergence diagnostic (quick and dirty)
  maxit <- c("maxit" = fit@control@maxit)
  # Update stk
  stk   <- transform(stk, harvest = harvest(fit), stock.n = stock.n(fit))
  return(list(stk = stk, converge = maxit))
}
```

**Control object for projections**  

The `fwd` method from `FLash` needs a control object, which is set by this function.

```{r, ctrlproj}
getCtrl <- function(values, quantity, years, it){
	dnms <- list(iter=1:it, year=years, c("min", "val", "max"))
	arr0 <- array(NA, dimnames=dnms, dim=unlist(lapply(dnms, length)))
	arr0[,,"val"] <- unlist(values)
	arr0 <- aperm(arr0, c(2,3,1))
	ctrl <- fwdControl(data.frame(year=years, quantity=quantity, val=NA))
	ctrl@trgtArray <- arr0
	ctrl
}
```

## MSE initialisation

The first year of the projection window is the intermediate year, for which, following the ICES WG timeline, already has a TAC. For this tutorial, the TAC in the final year of data is assumed to be the realised catch in `stk` for the same year, while the TAC in the intermediate year is set equal to the TAC in the final year of data. The `stk` object then needs to be projected through the intermediate year by applying the TAC in the intermediate year with `fwd`.

```{r, mseinit}
vy <- ac(iy:fy)
TAC <- FLQuant(NA, dimnames=list(TAC="all", year=c(dy,vy), iter=1:it))
TAC[,ac(dy)] <- catch(stk_a4a)[,ac(dy)]
TAC[,ac(iy)] <- TAC[,ac(dy)] #assume same TAC in the first intermediate year
ctrl <- getCtrl(c(TAC[,ac(iy)]), "catch", iy, it)
# Set up the operating model FLStock object
stk.om <- fwd(stk_a4a, control=ctrl, sr=srbh, sr.residuals = exp(srbh.res), sr.residuals.mult = TRUE)
```

## Start the MSE loop  

The MSE loop requires the following:

- observation error model to be applied to generate the index of abundance,
- the stock assessment (XSA) to be applied using this index to generate the management procedure stock object `stk.mp`,
- the resultant SSB estimate to be used in the HCR, along with the reference points obtained earlier, to derive target Fs for the TAC year (the year after the intermediate year), and
- the TAC associated with the target F to be calculated by applying `fwd` to the `stk.mp`.

The final step of the MSE loop is to apply the TAC to the operating model stock object, `stk.om`, by using `fwd`.
  
```{r, mseloop, results="hide"}
set.seed(231) # set seed to ensure comparability between different runs
for(i in vy[-length(vy)]){
  # set up simulations parameters
	ay <- an(i)
	cat(i, " > ")
  flush.console()
	vy0 <- 1:(ay-y0) # data years (positions vector) - one less than current year
	sqy <- (ay-y0-nsqy+1):(ay-y0) # status quo years (positions vector) - one less than current year
  # apply observation error
	oem <- o(stk.om, idx, i, vy0)
	stk.mp <- oem$stk
	idx.mp <- oem$idx
	idx <- oem$idx.om
  # perform assessment
  out.assess <- xsa(stk.mp, idx.mp)
  stk.mp <- out.assess$stk
  
  # apply ICES MSY-like Rule to obtain Ftrgt
  # (note this is not the ICES MSY rule, but is similar)
  flag <- ssb(stk.mp)[,ac(ay-1)]<Bpa
  Ftrgt <- ifelse(flag,ssb(stk.mp)[,ac(ay-1)]*Fmsy/Bpa,Fmsy) 
  # project the perceived stock to get the TAC for ay+1
  fsq.mp <- yearMeans(fbar(stk.mp)[,sqy]) # Use status quo years defined above
  ctrl <- getCtrl(c(fsq.mp, Ftrgt), "f", c(ay, ay+1), it)
  stk.mp <- stf(stk.mp, 2)
  gmean_rec <- c(exp(yearMeans(log(rec(stk.mp)))))
  stk.mp <- fwd(stk.mp, control=ctrl, sr=list(model="mean", params = FLPar(gmean_rec,iter=it)))
  TAC[,ac(ay+1)] <- catch(stk.mp)[,ac(ay+1)]
  # apply the TAC to the operating model stock
  ctrl <- getCtrl(c(TAC[,ac(ay+1)]), "catch", ay+1, it)
  stk.om <- fwd(stk.om, control=ctrl,sr=srbh, sr.residuals = exp(srbh.res), sr.residuals.mult = TRUE)
}
```

# PERFORMANCE STATISTICS

```{r, pstats}
#Some example performance statstics, but first isolate the projection period
stk.tmp<-window(stk.om,start=iy)
# annual probability of being below Blim
(risky<-iterSums(ssb(stk.tmp)/Blim<1)/it)
# mean probabiity of being below Blim in the first half of the projection period
mean(risky[,1:trunc(length(risky)/2)])
# ...and second half
mean(risky[,(trunc(length(risky)/2)+1):length(risky)])
# plot of SSB relative to Bmsy
boxplot(data~year,data=as.data.frame(ssb(stk.tmp)),main="SSB");abline(h=Bmsy,col="red")
# plot of landings relative to MSY yield
boxplot(data~year,data=as.data.frame(landings(stk.tmp)),main="Landings");abline(h=msy,col="red")
```

```{r fig1, fig.cap="Results for applying an ICES MSY-like rule, comparing the operating model to the management procedure"}
plot(FLStocks(stk.om=stk.om, stk.mp=stk.mp)) + theme(legend.position="top") + geom_vline(aes(xintercept=2009))
```

